---
title: "Exploring Fires"
subtitle: "My subtitle if needed"
author: 
  - Ellie Murray
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tinytable)
library(tidyverse)

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))
```


# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....



# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

```{r}
analysis_data |>
  mutate(occupancy_month = month(
    occupancy_date,
    label = TRUE,
    abbr = FALSE
  )) |>
  arrange(month(occupancy_date)) |>
  drop_na(occupied_beds) |>
  summarise(number_occupied = mean(occupied_beds),
            .by = occupancy_month) |>
  tt() |> 
  format_tt(escape = TRUE)
```

Overview text

## Measurement
	
Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.



Some of our data is of penguins (@fig-bills), from @palmerpenguins.

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)


Talk way more about it. 

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.








# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}



\newpage


# References


